---
title: "Lab9 Programming in R - Regular Expressions"
author: "Andrew Romasco"
date: "10/26/2020"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Matching Patterns 

```{r}
library(stringr)
library(tidyr)
library(tibble)
library(tidyverse)
```

### 14.3.1 Basic matches

The simplest patterns match exact strings:

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "an")
```

The next step up in complexity is `.`, which matches any character (except a newline):

```{r}
str_view(x, ".a.")
```

A way create regular expressions!

```{r}
# To create the regular expression, we need \\
dot <- "\\."
# But the expression itself only contains one:
writeLines(dot)

x <- c("abc", "a.c", "bef")
str_view(x, "a\\.c")
```

Examples of escaping from a string.

```{r}
x <- "a\\b"
writeLines(x)
str_view(x, "\\\\")
```

### Excercise

1. Explain why each of these strings don’t match a `\`: `"\"`, `"\\"`, `"\\\"`

  Each of these strings does not match because they each represent different things. The 
  first string would appear as just a \ string, whereas the second string would be        represented as "\" with the quotation marks included, whereas the third string is represented as "\\", and the fourth string is represented as "\\\".
  
### 14.3.2 Anchors

By default, regular expressions will match any part of a string. It’s often useful to anchor the regular expression so that it matches from the start or end of the string. You can use:

* `^` to match the start of the string.
* `$` to match the end of the string.

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
```

```{r}
str_view(x, "a$")
```

A way to remember is if you begin with power (`^`), you end up with money (`$`).

```{r}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
```

```{r}
str_view(x, "^apple$")
str_view(x, "$^$")

```

#### 14.3.2.1 Exercises

1. How would you match the literal string `"$^$"`?

  This should only be matched on the string "$^$" itself and not include any other strings that might include "$^$" or be used to detect other strings.
  
### Character classes and alternatives

There are a number of special patterns that match more than one character. You’ve already seen ., which matches any character apart from a newline. There are four other useful tools:

* `\d` : matches any digit.
* `\s` : matches any whitespace (e.g. space, tab, newline).
* `[abc]` : matches a, b, or c.
* `[^abc]` : matches anything except a, b, or c.

Remember, to create a regular expression containing `\d` or `\s`, you’ll need to escape the `\` for the string, so you’ll type `"\\d"` or `"\\s"`.


```{r}
# Look for a literal character that normally has special meaning in a regex
x <- c("abc", "a.c", "a*c", "a c")
str_view(x, "a[.]c")
```

```{r}
str_view(x, ".[*]c")
```

```{r}
str_view(x, "a[ ]")
```

```{r}
#To pick either one or the other
x <- c("grey", "gray")
str_view(x, "gr(e|a)y")
```

#### 14.3.3.1 Exercises

1. Create regular expressions to find all words that:
    * Start with a vowel.
    * That only contain consonants. (Hint: thinking about matching “not”-vowels.)
    * End with `ed`, but not with `eed`.
    * End with `ing` or `ise`.

```{r}
str_subset(stringr::words, "^[aeiou]")

# Same things as above but NO MATCHES
str_subset(stringr::words, "[aeiou]", negate=TRUE)

# Words ending with ed but no eed
str_subset(stringr::words, "[^e]ed$")

#Words ending with either ing or ise
str_subset(stringr::words, "i(ng|se)$")
```

### Repetition

The next step up in power involves controlling how many times a pattern matches:

* `?`: 0 or 1
* `+`: 1 or more
* `*`: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "CC?")
```

```{r}
str_view(x, "CC+")
```

```{r}
str_view(x, 'C[LX]+')
```

Note that the precedence of these operators is high, so you can write: `colou?r` to match either American or British spellings. That means most uses will need parentheses, like `bana(na)+`.

You can also specify the number of matches precisely:

* `{n}`: exactly n
* `{n,}`: n or more
* `{,m}`: at most m
* `{n,m}`: between n and m

```{r}
str_view(x, "C{2}")
str_view(x, "C{2,}")
```

Adding the ? will find the shortest version possible of the searched string

```{r}
str_view(x, 'C{2,3}?')
```

#### 14.3.4.1 Exercises

1. Describe the equivalents of ?, +, * in {m,n} form.

? maps to {0,1} which means that there is either 0 or 1 of that character. + maps to {1,} which means that it will match 1 or more than 1 instance of the character. * maps to {0,} which means 0 or more instances.

### Grouping and backreferences

Earlier, you learned about parentheses as a way to disambiguate complex expressions. Parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with backreferences, like `\1`, `\2` etc. For example, the following regular expression finds all fruits that have a repeated pair of letters.

```{r}
#The (..)\\1 finds 2 letters in series which match each other.
str_view(fruit, "(..)\\1", match = TRUE)

str_view(stringr::words, "(.)\1\1", match = TRUE)
```

#### 14.3.5.1 Exercises

1. Describe, in words, what these expressions will match:
    * `(.)\1\1
    This would be any character that appears three times in a row.
    * `"(.)(.)\\2\\1"`
    This is a pair of characters followed by those same characters but reversed.
    * `(..)\1`
    Two characters that are repeated.
    * `"(.).\\1.\\1"`
    Any character followed by any character followed by the 1st character followed by any character followed by the last character.
    * `"(.)(.)(.).*\\3\\2\\1"`
    Any three characters followed by zero or more characters of any kind followed by the same three characters but in reverse order.
    
## 14.4 Tools

Now that you’ve learned the basics of regular expressions, it’s time to learn how to apply them to real problems. In this section you’ll learn a wide array of stringr functions that let you:

* Determine which strings match a pattern.
* Find the positions of matches.
* Extract the content of matches.
* Replace matches with new values.
* Split a string based on a match.

### 14.4.1 Detect matches

* Note the below takes advantage of the built in `word` vector that is loaded with the `stringr` library.

To determine if a character vector matches a pattern, use str_detect(). It returns a logical vector the same length as the input:

```{r}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

Remember that when you use a logical vector in a numeric context, `FALSE` becomes 0 and `TRUE` becomes 1. That makes `sum()` and `mean()` useful if you want to answer questions about matches across a larger vector:

```{r}
# How many common words start with t?
sum(str_detect(words, "^t"))
```

```{r}
# What proportion of common words end with a vowel?
mean(str_detect(words, "[aeiou]$"))
```

When you have complex logical conditions (e.g. match a or b but not c unless d) it’s often easier to combine multiple `str_detect()` calls with logical operators, rather than trying to create a single regular expression. For example, here are two ways to find all words that don’t contain any vowels:

```{r}
# Find all words containing at least one vowel, and negate
no_vowels_1 <- !str_detect(words, "[aeiou]")
```

```{r}
# Find all words consisting only of consonants (non-vowels)
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)
```

The results are identical, but I think the first approach is significantly easier to understand. If your regular expression gets overly complicated, try breaking it up into smaller pieces, giving each piece a name, and then combining the pieces with logical operations.

A common use of `str_detect()` is to select the elements that match a pattern. You can do this with logical subsetting, or the convenient `str_subset()` wrapper:

```{r}
words[str_detect(words, "x$")]
```

```{r}
str_subset(words, "x$")
```

Typically, however, your strings will be one column of a data frame, and you’ll want to use filter instead:

```{r}
library(tibble)
df <- tibble(
  word = words, 
  i = seq_along(word)
)

df %>% 
  filter(str_detect(word, "x$"))
```

A variation on `str_detect()` is `str_count()`: rather than a simple yes or no, it tells you how many matches there are in a string:

```{r}
x <- c("apple", "banana", "pear")
str_count(x, "a")
```

```{r}
# On average, how many vowels per word?
mean(str_count(words, "[aeiou]"))
```

It’s natural to use `str_count()` with `mutate()`:
```{r}
df %>% 
  mutate(
    vowels = str_count(word, "[aeiou]"),
    consonants = str_count(word, "[^aeiou]")
  )
```

Note that matches never overlap. For example, in `"abababa"`, how many times will the pattern `"aba"` match? Regular expressions say two, not three:

```{r}
str_count("abababa", "aba")
```

```{r}
str_view_all("abababa", "aba")
```

#### 14.4.1.1 Exercises

1. For each of the following challenges, try solving it by using both a single regular expression, and a combination of multiple str_detect() calls.
    * Find all words that start or end with x.
    * Find all words that start with a vowel and end with a consonant.
    * Are there any words that contain at least one of each different vowel?
```{r}
#Either start with x, or end with x
words[str_detect(words, "^x|x$")]

#All words starting with a vowel, with anything inbetween, and then ending with smething that is not a vowel
str_subset(words, "^[aeiou].*[^aeiou]$")

#Contain one of each vowel at least
words[str_detect(words, "a") &
  str_detect(words, "e") &
  str_detect(words, "i") &
  str_detect(words, "o") &
  str_detect(words, "u")]
```

### Extract matches

To extract the actual text of a match, use `str_extract()`. To show that off, we’re going to need a more complicated example. I’m going to use the Harvard sentences, which were designed to test VOIP systems, but are also useful for practicing regexps. These are provided in `stringr::sentences`:

```{r}
length(sentences)
```

```{r}
head(sentences)
```

Imagine we want to find all sentences that contain a colour. We first create a vector of colour names, and then turn it into a single regular expression:

```{r}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match
```

Now we can select the sentences that contain a colour, and then extract the colour to figure out which one it is:

```{r}
has_colour <- str_subset(sentences, colour_match)
matches <- str_extract(has_colour, colour_match)
head(matches)
```
* The above example has an error. 
```{r}
has_colour
```
Can your spot the issue with the regular expression? FOUND IT! The error is that it takes the word red from reared, so this expression is finding colors where there are none.

Note that `str_extract()` only extracts the first match. We can see that most easily by first selecting all the sentences that have more than 1 match:

```{r}
more <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more, colour_match)
```

```{r}
str_extract(more, colour_match)
```

This is a common pattern for stringr functions, because working with a single match allows you to use much simpler data structures. To get all matches, use `str_extract_all()`. It returns a list:

```{r}
str_extract_all(more, colour_match)
```

You’ll learn more about lists in [lists](https://r4ds.had.co.nz/vectors.html#lists) and [iteration](https://r4ds.had.co.nz/vectors.html#iteration).

If you use `simplify = TRUE`, `str_extract_all()` will return a matrix with short matches expanded to the same length as the longest:

```{r}
str_extract_all(more, colour_match, simplify = TRUE)
```

```{r}
x <- c("a", "a b", "a b c")
str_extract_all(x, "[a-z]", simplify = TRUE)
```

####  14.4.2.1 Exercises

1. In the previous example, you might have noticed that the regular expression matched “flickered”, which is not a colour. Modify the regex to fix the problem.

```{r}
# \b indicates a word boundary. THis means only look for actual colors, not peices of words which happen to be colors.
colour_match2 <- str_c("\\b(", str_c(colours, collapse = "|"), ")\\b")
colour_match2

more2 <- sentences[str_count(sentences, colour_match) > 1]
str_view_all(more2, colour_match2, match = TRUE)
```

### Grouped matches

Earlier in this chapter we talked about the use of parentheses for clarifying precedence and for backreferences when matching. You can also use parentheses to extract parts of a complex match. For example, imagine we want to extract nouns from the sentences. As a heuristic, we’ll look for any word that comes after “a” or “the”. 

Defining a “word” in a regular expression is a little tricky, so here I use a simple approximation: a sequence of at least one character that isn’t a space.

```{r}
noun <- "(a|the) ([^ ]+)"
has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)
has_noun %>% 
  str_extract(noun)
```

`str_extract()` gives us the complete match; `str_match()` gives each individual component. Instead of a character vector, it returns a matrix, with one column for the complete match followed by one column for each group:

```{r}
has_noun %>% 
  str_match(noun)
```

(Unsurprisingly, our heuristic for detecting nouns is poor, and also picks up adjectives like smooth and parked.)

If your data is in a tibble, it’s often easier to use `tidyr::extract()`. It works like `str_match()` but requires you to name the matches, which are then placed in new columns:

```{r}
#library(tidyr)

tibble(sentence = sentences) %>% 
  tidyr::extract(
    sentence, c("article", "noun"), "(a|the) ([^ ]+)", 
    remove = FALSE
  )
```

Like `str_extract()`, if you want all matches for each string, you’ll need `str_match_all()`.

#### 14.4.3.1 Exercises

1. Find all words that come after a “number” like “one”, “two”, “three” etc. Pull out both the number and the word.

```{r}
#Finds a number specified and then any word
number <- "\\b(one|two|three|four|five|six|seven|eight|nine|ten) +(\\w+)"
#Extracts the number and the word
sentences[str_detect(sentences, number)] %>%
  str_extract(number)
```

### Replacing matches

`str_replace()` and `str_replace_all()` allow you to replace matches with new strings. The simplest use is to replace a pattern with a fixed string:

```{r}
#Replaces the first instance of a vowel
x <- c("apple", "pear", "banana")
str_replace(x, "[aeiou]", "-")
```

```{r}
#Replaces all instances of vowels
str_replace_all(x, "[aeiou]", "-")
```

With `str_replace_all()` you can perform multiple replacements by supplying a named vector:

```{r}
#NOW THIS IS USEFUL!
x <- c("1 house", "2 cars", "3 people")
str_replace_all(x, c("1" = "one", "2" = "two", "3" = "three"))
```

Instead of replacing with a fixed string you can use backreferences to insert components of the match. In the following code, I flip the order of the second and third words.

```{r}
sentences %>% 
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>% 
  head(5)
```


#### 14.4.4.1 Exercises

1. Replace all forward slashes in a string with backslashes.

```{r}
#Tough question, because it knows that \ is an escape character
x <- c("app/le", "p/ear", "bana/na")
str_replace_all(x, "/", "\\\\")

```

### Splitting

Use `str_split()` to split a string up into pieces. For example, we could split sentences into words:

```{r}
sentences %>%
  head(5) %>% 
  str_split(" ")
```
Because each component might contain a different number of pieces, this returns a list. If you’re working with a length-1 vector, the easiest thing is to just extract the first element of the list:
```{r}
"a|b|c|d" %>% 
  str_split("\\|") %>% 
  .[[1]]
```
Otherwise, like the other stringr functions that return a list, you can use `simplify = TRUE` to return a matrix:
```{r}
sentences %>%
  head(5) %>% 
  str_split(" ", simplify = TRUE)
```
You can also request a maximum number of pieces:

```{r}
fields <- c("Name: Hadley", "Country: NZ", "Age: 35")
fields %>% str_split(": ", n = 2, simplify = TRUE)
```

Instead of splitting up strings by patterns, you can also split up by character, line, sentence and word `boundary()`s:
```{r}
x <- "This is a sentence.  This is another sentence."
str_view_all(x, boundary("word"))
```

```{r}
str_split(x, " ")[[1]]
```

```{r}
str_split(x, boundary("word"))[[1]]
```

#### 14.4.5.1 Exercises

1. Split up a string like "apples, pears, and bananas" into individual components.

```{r}
# The answer in the key does not include the word and, but I am because it makes more sense to me like this based on what the question is asking.
str_split("apples, pears, and bananas", boundary("word"))[[1]]
```

### Stringi

I can't imagine a scenario where I might need it, but if I ever need more functions for strings Stringi is a much more comprehensive list of string functions than stringr.
