---
title: "Lab4A_Data_Wrangling"
author: "Andrew Romasco"
date: "9/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r Reading in Data}
report_03_11_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-11-2020.csv")) %>%
  rename(Country_Region = "Country/Region", Province_State = "Province/State")

report_09_13_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv"))

```

```{r Interactive Table}
#install.packages("DT")
library(DT)
datatable(report_03_11_2020)

datatable(report_09_13_2020)
```

## Exercises
1.)
```{r}
#str(report_03_11_2020)
#View(report_03_11_2020)

#str(report_09_13_2020)
#View(report_09_13_2020)
```
One of the differences between the two files is in how the column headers are named. In the file from 3_11_2020, there are a number of " / " used to seperate things like "Country/Region". In the newer files those forward slashes are replaced with underscores. In the new data, 9_13_2020, there are also new column headers such as: Active, Combined_Key, Incidence_Rate, and Case-Fatality_Ratio. In addition the latitude and longitude columns are now closer to the other columns which denote location such as the Country/Region column. Finally, there is a lot more data in the 9_13_2020 data set, 3955 lines, as opposed to the 3_11_2020 data set, 231 lines.

2.)
```{r}
total_report_03_11_2020 <- report_03_11_2020 %>%
  group_by(Country_Region) %>%
  summarize(
    Confirmed = sum(Confirmed),
    Deaths = sum(Deaths),
    Recovered = sum(Recovered))

head(total_report_03_11_2020)

total_report_09_13_2020 <- report_09_13_2020 %>%
  group_by(Country_Region) %>%
  summarize(
    Confirmed = sum(Confirmed),
    Deaths = sum(Deaths),
    Recovered = sum(Recovered),
    Active = sum(Active))

head(total_report_09_13_2020)
```

3.)
```{r}
US_report_03_11_2020 <- report_03_11_2020 %>%
    filter(Country_Region == "US") %>%
    group_by(Province_State) %>%
  summarize(
    Confirmed = sum(Confirmed),
    Deaths = sum(Deaths),
    Recovered = sum(Recovered))

head(US_report_03_11_2020)

US_report_09_13_2020 <- report_09_13_2020 %>%
    filter(Country_Region == "US") %>%
    group_by(Province_State) %>%
  summarize(
    Confirmed = sum(Confirmed),
    Deaths = sum(Deaths),
    Recovered = sum(Recovered),
    Active = sum(Active))

head(US_report_09_13_2020)
```

4.)
```{r}
short_US_report_03_11_2020 <- US_report_03_11_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

short_US_report_03_11_2020

short_US_report_09_13_2020 <- US_report_09_13_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

short_US_report_09_13_2020

```

5.)
```{r Geom Point Cases and Deaths}
#Top 10 Countries with confirmed Cases 03_11_2020
short_total_report_03_11_2020 <- total_report_03_11_2020 %>%
  arrange(desc(Confirmed)) %>% 
  slice(n = 1:10)

ggplot(data = short_total_report_03_11_2020, mapping = aes(x = Country_Region, y = Confirmed)) +
  aes(color = Country_Region) +
  geom_point(alpha = 1)

#Top 10 Countries with confirmed Cases 09_13_2020
short_total_report_09_13_2020 <- total_report_09_13_2020 %>%
  arrange(desc(Confirmed)) %>% 
  slice(n = 1:10)

ggplot(data = short_total_report_09_13_2020, mapping = aes(x = Country_Region, y = Confirmed)) +
  aes(color = Country_Region) +
  geom_point(alpha = 1)

#Top 10 Countries with confirmed Deaths 03_11_2020
short_total_report_03_11_2020 <- total_report_03_11_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

ggplot(data = short_total_report_03_11_2020, mapping = aes(x = Country_Region, y = Deaths)) +
  aes(color = Country_Region) +
  geom_point(alpha = 1)

#Top 10 Countries with confirmed Deaths 09_13_2020
short_total_report_09_13_2020 <- total_report_09_13_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

ggplot(data = short_total_report_09_13_2020, mapping = aes(x = Country_Region, y = Deaths)) +
  aes(color = Country_Region) +
  geom_point(alpha = 1)
```

6.)
```{r Geom Bar Top 10 Cases and Deaths US States}
#Top 10 States with confirmed Cases 03_11_2020
short_US_report_03_11_2020 <- US_report_03_11_2020 %>%
  arrange(desc(Confirmed)) %>% 
  slice(n = 1:10)

ggplot(data = short_US_report_03_11_2020, mapping = aes(x = Province_State, y = Confirmed)) +
  aes(color = Province_State) +
  geom_point(alpha = 1)

#Top 10 States with confirmed Cases 09_13_2020
short_US_report_09_13_2020 <- US_report_09_13_2020 %>%
  arrange(desc(Confirmed)) %>% 
  slice(n = 1:10)

ggplot(data = short_US_report_09_13_2020, mapping = aes(x = Province_State, y = Confirmed)) +
  aes(color = Province_State) +
  geom_point(alpha = 1)

#Top 10 States with confirmed Deaths 03_11_2020
short_US_report_03_11_2020 <- US_report_03_11_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

ggplot(data = short_US_report_03_11_2020, mapping = aes(x = Province_State, y = Deaths)) +
  aes(color = Province_State) +
  geom_point(alpha = 1)

#Top 10 Countries with confirmed Deaths 09_13_2020
short_US_report_09_13_2020 <- US_report_09_13_2020 %>%
  arrange(desc(Deaths)) %>% 
  slice(n = 1:10)

ggplot(data = short_US_report_09_13_2020, mapping = aes(x = Province_State, y = Deaths)) +
  aes(color = Province_State) +
  geom_point(alpha = 1)
```

