---
title: "lab3B_AVRR-Manipulating the Data"
author: "Andrew Romasco"
date: "9/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("tidyverse")
surveys <- read.csv("~/Andrew UMass Amherst Undergrad/Andrew Year 4 UMass/Fall 2020/Bio 597GE Genomics/genomics-course/data/portal_data_joined.csv")
```

## Selecting Columns

```{r}
#Select the 3 columns specified
#select(surveys, plot_id, species_id, weight)
#Select all but the columns specified
#select(surveys, -record_id, -species_id)
#To choose rows
#filter(surveys, year == 1995)
```

## Pipes

```{r}
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)
# OR use pipelines
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)

# CHALLENGE
surveys %>%
  filter(year < 1995) %>%
  select(year, sex, weight) %>%
  head()

## MUTATE
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
#Can make multiple new columns in one mutate statement
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2) %>%
  head()
# So you only see the first few rows
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
#Get rid of those NA's!
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()

#CHALLENGE
surveys_hindfoot_cm <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_cm = hindfoot_length / 10) %>%
    filter(hindfoot_cm < 3) %>%
    select(species_id, hindfoot_cm)
```

## Summarize and Group_by

group_by() is often used together with summarize(), which collapses each group into a single-row summary of that group. group_by() takes as arguments the column names that contain the categorical variables for which you want to calculate the summary statistics. So to compute the mean weight by sex:

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  head()

surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>% 
  tail()

surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  head()

#Use print to see even more rows
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight)) %>%
  print(n = 15)

#Multiple summaries at the same time
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  print(n = 15)
#Multiple summaries sorted with arrange
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight) %>%
  print(n = 15)
#Sorting by the mean weight but in descending order
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight)) %>%
  print(n = 15)
```

## Count
```{r}
#Count the number of rows for each sex
surveys %>%
    count(sex) 
#It can even sort if you want that
surveys %>%
    count(sex, sort = TRUE) 
#Counting a combination of factors
surveys %>%
  count(sex, species)  %>%
  print(n = 15)
#Now to organize our counts
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n)) %>%
  print(n = 15)

#CHALLENGE 1
surveys %>%
    count(plot_type) 
#CHALLENGE 2
surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species_id) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length),
        n = n()
    )
#CHALLENGE 3, and also arranged by year
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    filter(weight == max(weight)) %>%
    select(year, genus, species, weight) %>%
    arrange(year) %>%
  print(n = 15)
```

## Spread

Refer back to this function because it is amazing.

```{r}
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(plot_id, genus) %>%
  summarize(mean_weight = mean(weight))

str(surveys_gw)

#Spread out the genus column into a new column for each genus. Then set the mean weights as the rows.
surveys_spread <- surveys_gw %>%
  spread(key = genus, value = mean_weight)

str(surveys_spread)

surveys_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  head()
```

## Gather

The opposite of spread

```{r}
#CHALLENGE 1
surveys_spread_genera <- surveys %>%
  group_by(plot_id, year) %>%
  summarize(n_genera = n_distinct(genus)) %>%
  spread(year, n_genera)
#the surveys data frame with year as columns, plot_id as rows, and the number of genera per plot as the values
surveys_spread_genera

#CHALLENGE 2
surveys_spread_genera %>%
  gather("year", "n_genera", -plot_id) %>%
  print(n = 15)

#CHALLENGE 3
surveys_long <- surveys %>%
  gather("measurement", "value", hindfoot_length, weight)

#CHALLENGE 4
surveys_long %>%
  group_by(year, measurement, plot_type) %>%
  summarize(mean_value = mean(value, na.rm=TRUE)) %>%
  spread(measurement, mean_value) %>%
  print(n = 15)
```

## Exporting Data

```{r}
surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex

## Extract the most common species_id
species_counts <- surveys_complete %>%
    count(species_id) %>% 
    filter(n >= 50)

## Only keep the most common species
surveys_complete <- surveys_complete %>%
  filter(species_id %in% species_counts$species_id)

dim(surveys_complete)

#Saving the data
write_csv(surveys_complete, path = "~/Andrew UMass Amherst Undergrad/Andrew Year 4 UMass/Fall 2020/Bio 597GE Genomics/genomics-course/data/surveys_complete.csv")
```

